###############lines 0-200 DEG analysis. Lines 205 onwards analysis for figures

```{r}

library(DESeq2)
library(dplyr)
library(ggplot2)
library(ComplexHeatmap)
library(circlize)
library(sporkforlife)
library(clusterProfiler)
library(biomaRt)

all_samples.counts <- read.delim("/rds/projects/c/croftap-croftapcarcia/RNA_ATAC_JAN2025/RNAseq/count/pipline/results/featurecounts/all_samples.counts.txt", comment.char="#", row.names = 1)




```

```{r}

all_counts <- all_samples.counts

all_counts <- all_counts %>% dplyr::select(-c(Chr, Start,End, Strand, Length))    #remove non-essential columns

colnames(all_counts) <- c("C3", "X1", "X3", "C1", "C2", "X2")  #rename columns

##########reorder samples
all_counts <- all_counts[,c(2,6,3,4,5,1)]

#make meta data from count matrix
meta_data=colnames(all_counts)
meta_data<-as.data.frame(meta_data)

#add a condition column
meta_data$condition <- c(rep("X", 3), rep("C", 3))
colnames(meta_data)<-c("sample", "condition")

#make factors
meta_data$sample<-as.factor(meta_data$sample)
meta_data$condition<-as.factor(meta_data$condition)
```

```{r}
order <- c("X", "C")  #the order you want your conditions to appear

dds <- DESeqDataSetFromMatrix(countData = all_counts,
                              colData = meta_data,
                              design = ~ condition)

keep <- rowSums(counts(dds)) >10
dds <- dds[keep, ]
    
dds$condition <- factor(dds$condition, levels = c("X","C"))

dds <- DESeq(dds, test = "Wald")

res <- results(dds)
res

res <- results(dds, name="condition_C_vs_X")
res <- results(dds, contrast=c("condition","C","X"))    
    
resultsNames(dds)

resOrdered <- res[order(res$pvalue),]   
res05 <- results(dds, alpha=0.05)    
sum(res05$padj < 0.05, na.rm=TRUE)    
    
plotMA(res05, ylim=c(-2,2))    
plotCounts(dds, gene=which.min(res$padj), intgroup="condition")

d <- plotCounts(dds, gene=which.min(res$padj), intgroup="condition", 
                returnData=TRUE)
library("ggplot2")
ggplot(d, aes(x=condition, y=count)) + 
  geom_point(position=position_jitter(w=0.1,h=0)) + 
  scale_y_log10(breaks=c(25,100,400))

vsd <- vst(dds)
head(assay(vsd), 3)

select <- order(rowMeans(counts(dds,normalized=TRUE)),
                decreasing=TRUE)[1:20]
df <- as.data.frame(colData(dds)[,c("condition","sample")])
pheatmap(assay(vsd)[select,], cluster_rows=FALSE, show_rownames=FALSE,
         cluster_cols=FALSE, annotation_col=df)

sampleDists <- dist(t(assay(vsd)))

sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- paste(vsd$condition, vsd$type, sep="-")
colnames(sampleDistMatrix) <- NULL
library(RColorBrewer)
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
pheatmap(sampleDistMatrix,
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         col=colors)

plotPCA(vsd, intgroup=c("condition", "sample"))
library(DEGreport)
degs <- degComps(dds, combs = "condition",
                 contrast = list( c("condition", "X", "C") ))

DEGs <- data.frame(results(dds, contrast=c("condition", "X", "C"), alpha = 0.05))

DEGs$cluster <- "NO"
DEGs$cluster[DEGs$log2FoldChange > 0.05] <- "UP"
DEGs$cluster[DEGs$log2FoldChange < -0.05] <- "DOWN"



result <- process_deseq_PCA(counts = all_counts, 
                             meta_data = meta_data, 
                             relevel_condition = order, 
                             min_gene_count = 200)

dds <- result$dds
pca_plot <- result$pca_plot
pca_plot

```

```{r}


library(gsfisher)
annotation_gs <- fetchAnnotation(species="mm", ensembl_version=NULL, ensembl_host=NULL)



custom_colors <- c("ctrl" = "red", "CAR" = "blue")

#define order of how you want to plot the heatmap
custom_order <- c("ctrl", "CAR")


run_deseq_contrast_heatmap(dds = dds, 
                           meta_data = meta_data, 
                           targetvar = "condition", 
                           p_value_threshold = 0.001, 
                           log2fc_threshold = 0.25, 
                           condition_colors = custom_colors,
                           condition_order = custom_order,
                           go_analysis = F, 
                           organism = "mm", 
                           annotation_gs = annotation_gs)



results_all_f <- results_all %>% filter(padj < 0.05)


index <- match(results_all_f$gene, annotation_gs$ensembl_id)
results_all_f$gene_name <- annotation_gs$gene_name[index]

write.csv(results_all_f, "/rds/projects/c/croftap-croftapcarcia/RNA_ATAC_JAN2025/RNAseq/analysis/DEGs.csv")

```

    
results_all_f <- results_all %>% filter(padj < 0.05)


index <- match(results_all_f$gene, annotation_gs$ensembl_id)
results_all_f$gene_name <- annotation_gs$gene_name[index]
    
    
```

```{r}


scale_sub_vsd <- scale_sub_vsd %>% as.data.frame()


index <- match(rownames(scale_sub_vsd), annotation_gs$ensembl_id)
rownames(scale_sub_vsd) <- annotation_gs$gene_name[index]


scale_sub_vsd <- as.matrix(scale_sub_vsd)



annotation_col <- data.frame(
  Group = c("ctrl", "ctrl", "ctrl", "CAR", "CAR", "CAR") # Example groupings
)
rownames(annotation_col) <- colnames(data) # Match to column names of the data


Heatmap(scale_sub_vsd,
        name = "Expression",
        cluster_rows = T,
        cluster_columns = F,
        row_title_rot = 0,
        row_title_gp = gpar(fontsize = 10, fontface = "bold"),
        row_names_gp = gpar(fontsize = 9),
        column_names_gp = gpar(fontsize = 10),
        show_row_dend = T,
        show_column_dend = T,
        heatmap_legend_param = list(title = "Expression\nValue"), show_row_names = F, border = T)


```


############ FIgure 3f
vsd <- vst(dds, blind = T)
vsd_mat <- assay(vsd)
norm_expr_mat<- t(scale(t(vsd_mat)))

ensembl_ids <- rownames(norm_expr_mat)

# Match each Ensembl ID to its gene name
idx <- match(ensembl_ids, annotation_gs$ensembl_id)

# Replace rownames with gene names (keeping original if no match)
rownames(norm_expr_mat) <- ifelse(
  is.na(idx),
  ensembl_ids,  
  annotation_gs$gene_name[idx]
)

##########manually order genes to match ATAC order- repeat this for damage, homeostasis, repair and quiescence genes
gene_order <- c("Fscn1", "Septin5", "Htr2a", "C1qtnf3",
                "Tnfrsf9", "Cxcr6", "Fbn2", "Tnn",
                "Tnc", "Cxcl5", "Saa4")

# Make sure genes exist
gene_order <- gene_order[gene_order %in% rownames(norm_expr_mat)]

Heatmap(
  norm_expr_mat[gene_order, c(1,2,3,4,5,6)],
  name = "Z-score",
  border = TRUE,
  show_row_names = TRUE,
  cluster_rows = FALSE,      # IMPORTANT
  cluster_columns = FALSE,
  col = circlize::colorRamp2(c(-2, 0, 2),
                             c("blue", "white", "red"))
)

##########inflammation and complement
Heatmap(norm_expr_mat[c("Htr2a", "Saa4", "Septin5", "Cxcr6", "Fbn2", "Cxcl5", "Tnn", "Fscn1", "Tnfrsf9", "Tnc", "C1qtnf3"),c(1,2,3,4,5,6)], border=T, show_row_names = T, cluster_columns = F,c("blue", "white", "red"))+xlab(NULL)

##############damage
Heatmap(norm_expr_mat[c("Il1rn", "Il1rl1", "Pcsk5", "Adam12", "Tnfsf11", "Mmp8", "Mmp13", "Col8a2", "Col11a1"),c(1,2,3,4,5,6)], border=T, show_row_names = T, cluster_columns = F,c("blue", "white", "red"))+xlab(NULL)

###############homoeostasis
Heatmap(norm_expr_mat[c("Akap12", "Plau", "Itga6", "F13a1", "Btc"),c(1,2,3,4,5,6)], border=T, show_row_names = T, cluster_columns = F,c("blue", "white", "red"))+xlab(NULL)

#############repair
Heatmap(norm_expr_mat[c("Enho", "Pcolce2", "Prg4", "Cacng1", "Lepr", "Enpp6", "Tmigd1", "Mt1", "Mt2", "Megf10", "Cd24a"),c(1,2,3,4,5,6)], border=T, show_row_names = T, cluster_columns = F,c("blue", "white", "red"))+xlab(NULL)

##############T cell association
Heatmap(norm_expr_mat[c("H2-DMa", "Slfnlnc", "Ifi208", "Cxcl9", "H2-Ab1", "H2-Eb1", "H2-Aa"),c(1,2,3,4,5,6)], border=T, show_row_names = T, cluster_columns = F,c("blue", "white", "red"))+xlab(NULL)

###############quiescence
Heatmap(norm_expr_mat[c("Creb5", "Foxo1", "Sox5"),c(1,2,3,4,5,6)], border=T, show_row_names = T, cluster_columns = F,c("blue", "white", "red"))




```


\`\`\`{r}

#########pathway analysis - figure 3d&e

####using df_merge genes library(AnnotationDbi) library(org.Mm.eg.db)

DEGs <- data.frame(results(dds, contrast=c("condition", "X", "C"), alpha = 0.05))

DEGs$cluster <- "NO"
DEGs$cluster[DEGs$log2FoldChange > 0.05] <- "UP"
DEGs$cluster[DEGs$log2FoldChange < -0.05] <- "DOWN"

library(biomaRt)
library(gsfisher)

annotation_gs <- fetchAnnotation(species="mm", ensembl_version=NULL, ensembl_host=NULL)

ensembl <- useMart("ensembl", dataset = "mmusculus_gene_ensembl")
gene_ids <- rownames(DEGs)
gene_map <- getBM(
  attributes = c("ensembl_gene_id", "mgi_symbol"),
  filters = "ensembl_gene_id",
  values = gene_ids,
  mart = ensembl
)

DEGs$ensembl_gene_id <- rownames(DEGs)
df_merged <- merge(gene_map, DEGs, by = "ensembl_gene_id")

df_mergedres \<- subset(df_merged, padj \< 0.05 & abs(log2FoldChange) \> 1) df_mergedX \<- subset(df_mergedres, padj \< 0.05 & cluster == "UP") df_mergedC \<- subset(df_mergedres, padj \< 0.05 & cluster == "DOWN")

library(clusterProfiler)

\`\` egoX \<- enrichGO(gene = df_mergedX\$gene_id, OrgDb = org.Mm.eg.db, keyType = "ENTREZID", ont = "BP", pAdjustMethod = "BH", qvalueCutoff = 0.2, readable = TRUE)

#################Figure 3d
dotplot(egoX, showCategory = c("secretion by tissue", "acute inflammatory response", "humoral immune response", "complement activation", "regulation of complement activation", "complement activation, classical pathway", "learning or memory", "memory", "production of molecular mediator of immune response", "cell aggregation", "acute inflammatory response", "phagocytosis, recognition")) + ggtitle("GO Terms - non-treated")

egoC \<- enrichGO(gene = df_mergedX\$gene_id, OrgDb = org.Mm.eg.db, keyType = "ENTREZID", ont = "BP", pAdjustMethod = "BH", qvalueCutoff = 0.2, readable = TRUE)

###################Figure 3e
dotplot(egoC, showCategory = c("homeostasis of number of cells", "myeloid cell homeostasis", "B cell mediated immunity", "response to type II interferon", "homotypic cell-cell adhesion", "B cell homeostasis", "response to vitamin", "antigen processing and presentation of exogenous antigen")) + ggtitle("GO Terms - non-treated")
